WIDS 2024 Final Project

### 2024-01-11 Code/R FIle.qmd
worked on models

in the house data set, we can see a significant amount of NA values in the runoff column

@@

```{r}
per_collective_wage_data <- full_state_data |> # This object will be used to create the graph and also in the data subset
  filter(!is.na(regions)) |> #This just removes DC
  filter(year <= 2020 & year >= 1984) |>  #wage data only exists before 2020
  mutate(
    map_color = case_when(
      state_abbreviation == "CA" ~ "CA", # States we would like to highlight. first "" is the state abb, second "" is what will appear in legend
      state_abbreviation == "AK" ~ "AK",
      state_abbreviation == "HI" ~ "HI",
      state_abbreviation == "FL" ~ "FL",
      .default = NA
    )
  ) 
per_collective_wage_data |> # use created table here
  ggplot(aes(x = year)) +
  geom_line(aes(y = percent_collective_bargain, group = state_abbreviation),
            color = "grey75", alpha =0.6) + # using group means the groupings wont appear in legend
  geom_line(data = subset(per_collective_wage_data, # use created object here
                          !is.na(map_color)), #subset used to keep out non highlighted states
            aes(y = percent_collective_bargain, color = map_color),
            linewidth = .75) +
  geom_line(data = subset(full_ntl_data, year >= 1984),
            aes(y = percent_collective_bargain), color = "orange", linewidth = 1.5) +
  annotate("text", x=1988, y=.17, label= "Federal Avg", color = "orange", size = 5) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(per_collective_wage_data$year), max(per_collective_wage_data$year), by = 4)) + 
  labs(title = "States by % of Labor Market Covered By a Collective Bargaining Agreement",
       x = "Year",
       y = "% of Labor Market in a Collective Bargaining Agreement")

@@

```{r}
state_minimum_wage_data <- full_state_data |> # This object will be used to create the graph and also in the data subset
  filter(!is.na(regions)) |> #This just removes DC
  filter(year <=2020 & year >=1984) |>  #wage data only exists before 2020
  mutate(
    map_color = case_when(
      state_abbreviation == "CA" ~ "CA", # States we would like to highlight. first "" is the state abb, second "" is what will appear in legend
      state_abbreviation == "AK" ~ "AK",
      state_abbreviation == "OK" ~ "OK",
      state_abbreviation == "FL" ~ "FL",
      .default = NA
    )
  ) 
state_minimum_wage_data |> # use created table here
  ggplot(aes(x = year)) +
  geom_line(aes(y = state_min_wage_2020_adj, group = state_abbreviation),
            color = "grey75", alpha =0.6) + # using group means the groupings wont appear in legend
  geom_line(data = subset(state_minimum_wage_data, # use created object here
                          !is.na(map_color)), #subset used to keep out non highlighted states
            aes(y = state_min_wage_2020_adj, color = map_color),
            linewidth = .75) +
  geom_line(aes(y = fed_min_wage_2020_adj), color = "red", linewidth = .75) + # federal minimum wage
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(state_minimum_wage_data$year), max(state_minimum_wage_data$year), by = 4)) + 
  labs(title = "State Minimum Wage by Year (Adjusted for 2020 $)",
       subtitle = "For states with an adjusted effective minimum wage above the federal level",
       x = "Year",
       y = "Minimum Wage (2020 USD)")

@@

```{r}
effective_minimum_wage_data <- full_state_data |> 
  filter(!is.na(regions)) |> 
  filter(effective_min_wage_2020_adj != fed_min_wage_2020_adj) |> 
  filter(year <=2020 & year >=1984) |>
  mutate(
    map_color = case_when(
      state_abbreviation == "CA" ~ "CA",
      state_abbreviation == "AK" ~ "AK",
      state_abbreviation == "CT" ~ "CT",
      state_abbreviation == "HI" ~ "HI",
      .default = NA
    )
  ) 
effective_minimum_wage_data |> 
  ggplot(aes(x = year)) +
  geom_line(aes(y = effective_min_wage_2020_adj, group = state_abbreviation),
            color = "grey75", alpha =0.6) +
  geom_line(data = subset(effective_minimum_wage_data, !is.na(map_color)),
            aes(y = effective_min_wage_2020_adj, color = map_color),
            linewidth = .75) +
  geom_line(aes(y = fed_min_wage_2020_adj), color = "red", linewidth = .75) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(effective_minimum_wage_data$year), max(effective_minimum_wage_data$year), by = 4)) + 
  labs(title = "State Effective Minimum Wage by Year (Adjusted for 2020 $)",
       subtitle = "For states with an adjusted effective minimum wage above the federal level",
       x = "Year",
       y = "Effective minimum wage (2020 USD)")

@@

```{r}
full_ntl_data |> 
  filter(year >= 1984) |> 
  ggplot(aes(x = year)) +
  geom_line(aes(y = mean_income_2020_adj), color = "red") +
  geom_line(aes(y = median_income_2020_adj), color = "blue") +
  theme_minimal() +
  labs(
    title = "Mean and Median income in the United States Between 1984 and 2019",
    subtitle = "Using incomes adjusted to 2020 rates",
    x = "Year",
    y = "Income in USD (Adjusted for Inflation to 2020 Rates"
  )
```

```{r}
full_ntl_data |> 
  filter(year >= 1984) |> 
  ggplot(aes(x = year)) +
  geom_line(aes(y = mean_income), color = "red") +
  geom_line(aes(y = median_income), color = "blue") +
  theme_minimal() +
  labs(
    title = "Mean and Median income in the United States Between 1984 and 2019",
    subtitle = "Using incomes adjusted to 2020 rates",
    x = "Year",
    y = "Income in USD (Adjusted for Inflation to 2020 Rates"
  )
```

@@

rm(map_data)
```
We can see that, especially in the northern-midwest states, that the country has grown to be more republican in he last 30 years. However, in the last 10 years this has begun to shift again. During this whole time, it appears that New England has become more democrat.

@@

This graph is incredibly interesting. First, a quick view of this grpah shows the exact opposite of our hypothesis. Looking at each decade we can see collective bargaining rates decrease and wage increase. Then, in each decade that relationship continues. Still, we can see that in every decade, a collective bargaining agreement will lead to an increase in wages for a worker over a non-collective-bargaining-jobs.

@@

```{r}
full_state_data |> 
  filter(year >= 1984) |> 
  ggplot(aes(x = percent_collective_bargain, y = party_state_average)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  facet_wrap(~decade) +
  labs(
    title = "Percent collective Bargaining vs Proportion of House Delegation that is Democratic",
    x = "Percent of workers covered by Collective Bargaining",
    y = "Proportion of House Delegation that is Democration"
  )
```

This graph demonstrates how political collective bargaining has become over the last 40 years. In the 1980s, there was almost no relationship between collective bargaining agreements and party affiliation. However, by 2020, there is now a significant bias

```{r}
full_state_data |> 
  filter(year <2020 & year >= 1984) |> 
  filter(year %% 4==0) |>
  ggplot(aes(x = percent_collective_bargain, y = party_state_average)) +
  geom_jitter(aes(color = regions), alpha = 0.5, width = 0.1, height = 0.1) +
  geom_smooth(method = "lm", color = "black") +
  facet_wrap(~decade) +
  labs(
    title = "Percent collective Bargaining vs Proportion of House Delegation that is Democratic",
    x = "Percent of workers covered by Collective Bargaining",
    y = "Proportion of House Delegation that is Democration"
  )
```

@@

>>>>

In the house data set, we can see a significant amount of NA values in the runoff column

@@

```{r}
per_collective_wage_data <- full_state_data |> # This object will be used to create the graph and also in the data subset
  filter(!is.na(regions)) |> #This just removes DC
  filter(year <= 2020 & year >= 1984) |>  #wage data only exists before 2020
  mutate(
    map_color = case_when(
      state_abbreviation == "CA" ~ "CA", # States we would like to highlight. first "" is the state abb, second "" is what will appear in legend
      state_abbreviation == "AK" ~ "AK",
      state_abbreviation == "HI" ~ "HI",
      state_abbreviation == "FL" ~ "FL",
      .default = NA
    )
  ) 
per_collective_wage_data |> # use created table here
  ggplot(aes(x = year)) +
  geom_line(aes(y = percent_collective_bargain, group = state_abbreviation),
            color = "grey75", alpha =0.6) + # using group means the groupings wont appear in legend
  geom_line(data = subset(per_collective_wage_data, # use created object here
                          !is.na(map_color)), #subset used to keep out non highlighted states
            aes(y = percent_collective_bargain, color = map_color),
            linewidth = .75) +
  geom_line(data = subset(full_ntl_data, year >= 1984),
            aes(y = percent_collective_bargain), color = "orange", linewidth = 1.5) +
  annotate("text", x=1988, y=.17, label= "Federal Avg", color = "orange", size = 5) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(per_collective_wage_data$year), max(per_collective_wage_data$year), by = 4)) + 
  labs(title = "States by % of Labor Market Covered By a Collective Bargaining Agreement",
       x = "Year",
       y = "% of Labor Market in a Collective Bargaining Agreement",
       color = "Selected States")

@@

```{r}
state_minimum_wage_data <- full_state_data |> # This object will be used to create the graph and also in the data subset
  filter(!is.na(regions)) |> #This just removes DC
  filter(year <=2020 & year >=1984) |>  #wage data only exists before 2020
  mutate(
    map_color = case_when(
      state_abbreviation == "CA" ~ "CA", # States we would like to highlight. first "" is the state abb, second "" is what will appear in legend
      state_abbreviation == "AK" ~ "AK",
      state_abbreviation == "OK" ~ "OK",
      state_abbreviation == "FL" ~ "FL",
      .default = NA
    )
  ) 
state_minimum_wage_data |> # use created table here
  ggplot(aes(x = year)) +
  geom_line(aes(y = state_min_wage_2020_adj, group = state_abbreviation),
            color = "grey75", alpha =0.6) + # using group means the groupings wont appear in legend
  geom_line(data = subset(state_minimum_wage_data, # use created object here
                          !is.na(map_color)), #subset used to keep out non highlighted states
            aes(y = state_min_wage_2020_adj, color = map_color),
            linewidth = .75) +
  geom_line(aes(y = fed_min_wage_2020_adj), color = "red", linewidth = .75) + # federal minimum wage
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(state_minimum_wage_data$year), max(state_minimum_wage_data$year), by = 4)) + 
  labs(title = "State Minimum Wage by Year (Adjusted for 2020 $)",
       subtitle = "For states with an adjusted effective minimum wage above the federal level",
       x = "Year",
       y = "Minimum Wage (2020 USD)",
       color = "Selected States")

@@

```{r}
effective_minimum_wage_data <- full_state_data |> 
  filter(!is.na(regions)) |> 
  filter(effective_min_wage_2020_adj != fed_min_wage_2020_adj) |> 
  filter(year <=2020 & year >=1984) |>
  mutate(
    map_color = case_when(
      state_abbreviation == "CA" ~ "CA",
      state_abbreviation == "AK" ~ "AK",
      state_abbreviation == "CT" ~ "CT",
      state_abbreviation == "HI" ~ "HI",
      .default = NA
    )
  ) 
effective_minimum_wage_data |> 
  ggplot(aes(x = year)) +
  geom_line(aes(y = effective_min_wage_2020_adj, group = state_abbreviation),
            color = "grey75", alpha =0.6) +
  geom_line(data = subset(effective_minimum_wage_data, !is.na(map_color)),
            aes(y = effective_min_wage_2020_adj, color = map_color),
            linewidth = .75) +
  geom_line(aes(y = fed_min_wage_2020_adj), color = "red", linewidth = .75) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(effective_minimum_wage_data$year), max(effective_minimum_wage_data$year), by = 4)) + 
  labs(title = "State Effective Minimum Wage by Year (Adjusted for 2020 $)",
       subtitle = "For states with an adjusted effective minimum wage above the federal level",
       x = "Year",
       y = "Effective minimum wage (2020 USD)",
       color = "Selected States")

@@

```{r}
full_ntl_data |> 
  filter(year >= 1984) |> 
  ggplot(aes(x = year)) +
  geom_line(aes(y = mean_income_2020_adj), color = "red") +
  geom_line(aes(y = median_income_2020_adj), color = "blue") +
  theme_minimal() +
  labs(
    title = "Mean and Median Income in the United States Between 1984 and 2019",
    subtitle = "Using mean (red) and median (blue) incomes adjusted to 2020 rates",
    x = "Year",
    y = "Income in USD (Adjusted for Inflation to 2020 Rates"
  )
```

```{r}
full_ntl_data |> 
  filter(year >= 1984) |> 
  ggplot(aes(x = year)) +
  geom_line(aes(y = mean_income), color = "red") +
  geom_line(aes(y = median_income), color = "blue") +
  theme_minimal() +
  labs(
    title = "Mean and Median Income in the United States Between 1984 and 2019",
    subtitle = "Using mean (red) and median (blue) incomes adjusted to 2020 rates",
    x = "Year",
    y = "Income in USD (Adjusted for Inflation to 2020 Rates)"
  )
```

@@

rm(map_data)
```
We can see that, especially in the northern-midwest states, that the country has grown to be more republican in the last 30 years. However, in the last 10 years this has begun to shift again. During this whole time, it appears that New England has become more democrat.

@@

This graph is incredibly interesting. First, a quick view of this graph shows the exact opposite of our hypothesis. Looking at each decade we can see collective bargaining rates decrease and wage increase. Then, in each decade that relationship continues. Still, we can see that in every decade, a collective bargaining agreement will lead to an increase in wages for a worker over a non-collective-bargaining-jobs.

@@

```{r}
full_state_data |> 
  filter(year >= 1984) |> 
  ggplot(aes(x = percent_collective_bargain, y = party_state_average)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  facet_wrap(~decade) +
  labs(
    title = "Percent Collective Bargaining vs Proportion of House Delegation that is Democratic",
    x = "Percent of Workers Covered by Collective Bargaining",
    y = "Proportion of House Delegation that is Democratic"
  )
```

This graph demonstrates how political collective bargaining has become over the last 40 years. In the 1980s, there was almost no relationship between collective bargaining agreements and party affiliation. However, by 2020, there is now a significant bias.

```{r}
full_state_data |> 
  filter(year <2020 & year >= 1984) |> 
  filter(year %% 4==0) |>
  ggplot(aes(x = percent_collective_bargain, y = party_state_average)) +
  geom_jitter(aes(color = regions), alpha = 0.5, width = 0.1, height = 0.1) +
  geom_smooth(method = "lm", color = "black") +
  facet_wrap(~decade) +
  labs(
    title = "Percent Collective Bargaining vs Proportion of House Delegation that is Democratic",
    x = "Percent of Workers Covered by Collective Bargaining",
    y = "Proportion of House Delegation that is Democratic",
    color = "Region"
  )
```

This graph demonstrates the same as the previous graph, but with a linear regression line for each region.

```{r}
full_state_data |> 
  filter(year <2020 & year >= 1984) |> 
  filter(year %% 4==0) |>
  ggplot(aes(x = percent_collective_bargain, y = party_state_average)) +
  geom_jitter(aes(color = regions), alpha = 0.1, width = 0.1, height = 0.1) +
  geom_smooth(method = "lm", aes(group = regions, color = regions), se = FALSE) +
  facet_wrap(~decade) +
  labs(
    title = "Percent Collective Bargaining vs Proportion of House Delegation that is Democratic",
    x = "Percent of Workers Covered by Collective Bargaining",
    y = "Proportion of House Delegation that is Democratic",
    color = "Region"
  )
```

@@

T-test between union and non-union wages
```{r}
# 1984 to 2019
t.test(full_ntl_data$union_wage_2020_adj, full_ntl_data$nonunion_wage_2020_adj)

# 1980s
full_ntl_data_1980s <- full_ntl_data |>
  filter(year >= 1984 & year < 1990)
t.test(full_ntl_data_1980s$union_wage_2020_adj, full_ntl_data_1980s$nonunion_wage_2020_adj)

# 1990s
full_ntl_data_1990s <- full_ntl_data |>
  filter(year >= 1990 & year < 2000)
t.test(full_ntl_data_1990s$union_wage_2020_adj, full_ntl_data_1990s$nonunion_wage_2020_adj)

# 2000s
full_ntl_data_2000s <- full_ntl_data |>
  filter(year >= 2000 & year < 2010)
t.test(full_ntl_data_2000s$union_wage_2020_adj, full_ntl_data_2000s$nonunion_wage_2020_adj)

# 2010s
full_ntl_data_2010s <- full_ntl_data |>
  filter(year >= 2010 & year < 2020)
t.test(full_ntl_data_2010s$union_wage_2020_adj, full_ntl_data_2010s$nonunion_wage_2020_adj)
```

T test between median and mean incomes
```{r}
# 1984 to 2019
t.test(full_ntl_data$median_income_2020_adj, full_ntl_data$mean_income_2020_adj)

# 1980s
full_ntl_data_1980s <- full_ntl_data |>
  filter(year >= 1984 & year < 1990)
t.test(full_ntl_data_1980s$median_income_2020_adj, full_ntl_data_1980s$mean_income_2020_adj)

# 1990s
full_ntl_data_1990s <- full_ntl_data |>
  filter(year >= 1990 & year < 2000)
t.test(full_ntl_data_1990s$median_income_2020_adj, full_ntl_data_1990s$mean_income_2020_adj)

# 2000s
full_ntl_data_2000s <- full_ntl_data |>
  filter(year >= 2000 & year < 2010)
t.test(full_ntl_data_2000s$median_income_2020_adj, full_ntl_data_2000s$mean_income_2020_adj)

# 2010s
full_ntl_data_2010s <- full_ntl_data |>
  filter(year >= 2010 & year < 2020)
t.test(full_ntl_data_2010s$median_income_2020_adj, full_ntl_data_2010s$mean_income_2020_adj)
```

Regression between percent union and party state average
```{r}
# 1984 to 2019
summary(lm(full_state_data$party_state_average~full_state_data$percent_union))

# 1980s
full_state_data_1980s <- full_state_data |>
  filter(year >= 1984 & year < 1990)
summary(lm(full_state_data_1980s$party_state_average~full_state_data_1980s$percent_union))

# 1990s
full_state_data_1990s <- full_state_data |>
  filter(year >= 1990 & year < 2000)
summary(lm(full_state_data_1990s$party_state_average~full_state_data_1990s$percent_union))

# 20000s
full_state_data_2000s <- full_state_data |>
  filter(year >= 2000 & year < 2010)
summary(lm(full_state_data_2000s$party_state_average~full_state_data_2000s$percent_union))

# 2010s
full_state_data_2010s <- full_state_data |>
  filter(year >= 2010 & year < 2020)
summary(lm(full_state_data_2010s$party_state_average~full_state_data_2010s$percent_union))
```
